---
title: "P8105_HW2_DI2251"
author: "di2251"
date: "2026-10-03"
output: github_document
---
```{r setup, echo=FALSE, message=FALSE}
library(haven)
library(tidyverse)
library(readxl)
library(ggridges)
```

### Problem 1

```{r data_import_q1}
NYC_subway = 
  read_csv(file = "data/NYC Transit Subway Data.csv", na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude, station_longitude, 
         starts_with("route"), entry, vending, entrance_type, ada) |>
  mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))

NYC_subway
```

**Data-set description:**
This data-set contains information related to each entrance and exit for each subway station in New York City. As a part of the initial data cleaning, the variable names of the imported data were automatically reformatted to a standardized format. The 19 variables or columns relevant to the problem were selected out of the 32 total variables (columns) in the original data-set. These selected variables contain information on the subway line, station name, station latitude/longitude, route number, entry, vending, entrance type, and ADA compliance. There are a total of 1,868 rows or observations in the data-set. In addition, the entry variable was converted from a character variable of `YES` vs `NO` to a logical variable. However, this data is not tidy since the lines that run or do not run through each of routes 1 to 11 are independently displayed the information can be collapsed into a character variable listing which lines travel along which routes rather than having them all listed separately. 

```{r}
NYC_subway |> distinct(line, station_name)
```

**How many distinct stations are there?** *Note: stations are identified both by name and by line*

There are 465 distinct stations. 

```{r}
filter(NYC_subway, ada == TRUE)
```

**How many stations are ADA compliant?**

468 stations are ADA compliant. 

```{r}
filter(NYC_subway, vending == "NO")

filter(NYC_subway, vending == "NO" & entry == TRUE)
```

**What proportion of station entrances/exits without vending allow entrance?**

69 subway stations out of 183 subway stations without vending allow entrance. 

```{r}
NYC_subway_route = NYC_subway |>
  mutate(across(route8:route11, as.character)) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) |>
  drop_na(route_name)

NYC_subway_route |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

NYC_subway_route |>
  filter(route_name == "A", ada == TRUE) |>
  distinct (line, station_name) |>
  nrow()
```

**How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?**

60 distinct stations serve the A train and of those, 17 are ADA compliant. 

### Problem 2

```{r excel_info_q2}
excel_sheets("data/Trash Wheel Collection Data 2026.xlsx")
```
Import `Mr. Trash Wheel` sheet. 

```{r data_import_q2.1, message=FALSE}
mr_trash = read_excel(
  "data/Trash Wheel Collection Data 2026.xlsx",
  sheet = "Mr. Trash Wheel",
  range = "A2:N742",
  na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    sports_balls = as.integer(round(sports_balls))
  )
```

Import `Professor Trash Wheel` and `Gwynnda the Good Wheel of the W` sheets.

```{r data_import_q2.2, message=FALSE}
prof_trash = read_excel(
  "data/Trash Wheel Collection Data 2026.xlsx",
  sheet = "Professor Trash Wheel",
  range = "A2:M139",
  na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  drop_na(dumpster)

gwynnda_df = read_excel(
  "data/Trash Wheel Collection Data 2026.xlsx",
  sheet = "Gwynnda the Good Wheel of the W",
  range = "A2:L394",
  na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  drop_na(dumpster)
```

Label each trash wheel before merging. Identified variables stored in different format and rectified differences.  

```{r label_df}
mr_trash = mr_trash |>
  mutate(trash_wheel = "Mr. Trash Wheel", year = as.numeric(year))

mr_trash

prof_trash = prof_trash |>
  mutate(trash_wheel = "Professor Trash Wheel", plastic_bottles = as.numeric(plastic_bottles))

prof_trash

gwynnda_df = gwynnda_df |>
  mutate(trash_wheel = "Gwynnda the Good Wheel of the W")

gwynnda_df
```

Merge all datasets.

```{r}
all_trash = bind_rows(mr_trash, prof_trash, gwynnda_df)
```

```{r trash_summary, echo = FALSE}
prof_total_weight = all_trash |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |>
  sum(na.rm = TRUE)

gwynnda_june2022_butts = all_trash |>
  filter(trash_wheel == "Gwynnda the Good Wheel of the W", month == "June", year == "2022") |>
  pull(cigarette_butts) |>
  sum(na.rm = TRUE)
```

The combined dataset contains `r nrow(all_trash)` observations from `r n_distinct(all_trash$trash_wheel)` trash wheels, each row representing one dumpster of collected trash. Key variables include `r names(all_trash)`.

The total weight of trash collected by Professor Trash Wheel was `r prof_total_weight` tons. In June 2022, Gwynnda collected `r format(gwynnda_june2022_butts, big.mark = ",")` cigarette butts. 

### Problem 3

```{r data import_q3}
MCI_baseline = read_csv(
  file = "data/data_mci/MCI_baseline.csv",
  na = c("", "NA", ".")
)

MCI_baseline = janitor::clean_names(MCI_baseline)

MCI_amyloid = read_csv(
  file = "data/data_mci/mci_amyloid.csv",
  na = c("", "NA", ".")
)

MCI_amyloid = janitor::clean_names(MCI_amyloid)
```






