---
title: "P8105_HW2_DI2251"
author: "di2251"
date: "2026-10-03"
output: github_document
---
```{r setup, echo=FALSE, message=FALSE}
library(haven)
library(tidyverse)
library(readxl)
library(ggridges)
```

### Problem 1

```{r data_import_q1}
NYC_subway = 
  read_csv(file = "data/NYC Transit Subway Data.csv", na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude, station_longitude, 
         starts_with("route"), entry, vending, entrance_type, ada) |>
  mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))

NYC_subway
```

**Data-set description:**
This data-set contains information related to each entrance and exit for each subway station in New York City. As a part of the initial data cleaning, the variable names of the imported data were automatically reformatted to a standardized format. The 19 variables or columns relevant to the problem were selected out of the 32 total variables (columns) in the original data-set. These selected variables contain information on the subway line, station name, station latitude/longitude, route number, entry, vending, entrance type, and ADA compliance. There are a total of 1,868 rows or observations in the data-set. In addition, the entry variable was converted from a character variable of `YES` vs `NO` to a logical variable. However, this data is not tidy since the lines that run or do not run through each of routes 1 to 11 are independently displayed the information can be collapsed into a character variable listing which lines travel along which routes rather than having them all listed separately. 

```{r}
NYC_subway |> distinct(line, station_name)
```

**How many distinct stations are there?** *Note: stations are identified both by name and by line*

There are 465 distinct stations. 

```{r}
filter(NYC_subway, ada == TRUE)
```

**How many stations are ADA compliant?**

468 stations are ADA compliant. 

```{r}
filter(NYC_subway, vending == "NO")

filter(NYC_subway, vending == "NO" & entry == TRUE)
```

**What proportion of station entrances/exits without vending allow entrance?**

69 subway stations out of 183 subway stations without vending allow entrance. 

### Problem 2

```{r data import_q2}
Trash_Wheel = read_excel(
  "data/Trash Wheel Collection Data 2026.xlsx",
  na = c("", "NA", ".")
)

Trash_Wheel = janitor::clean_names(Trash_Wheel)
```



### Problem 3

```{r data import_q3}
MCI_baseline = read_csv(
  file = "data/data_mci/MCI_baseline.csv",
  na = c("", "NA", ".")
)

MCI_baseline = janitor::clean_names(MCI_baseline)

MCI_amyloid = read_csv(
  file = "data/data_mci/mci_amyloid.csv",
  na = c("", "NA", ".")
)

MCI_amyloid = janitor::clean_names(MCI_amyloid)
```






